Rapid Epidemiological Analysis of Comorbidities and Treatments in COVID-19 in Scotland (REACT-SCOT): a case-control study based on the Scottish population [**Draft, not to be quoted**]

### 2020-05-20 casecontrol.Rmd
updated for pharmacoepi, incidencemortality.R refactored as separate file

_Background_-- Why some people infected with SARS-CoV-2 develop severe COVID-19 disease and others do not is poorly understood.  Here we  used sociodemographic data and extensive previous hospitalisation and prescribing history among all people who have developed severe COVID-19 disease in Scotland compared to the background population to understand factors associated with severity.

_Methods_ -- We linked the records of those with a positive nucleic acid test  for SARS-CoV-2 in the national database that captures virology testing in all NHS laboratories, to  hospital admissions, intensive care and mortality data. Severe COVID-19 was defined as disease requiring intensive care support or associated with death within 28 days of a positive test. We selected 10 sex, age and general practice based controls per test-positive case. For all cases and controls we extracted the International Classification of Disease (ICD)-10 codes from the previous five years of hospital discharge records  and the British National Formulary (BNF)  drug codes of  encashed  primary care issued prescriptions in the past year. We evaluated the risk of severe COVID-19 associated with  a pre-defined  set of  conditions using conditional logistic regression. We examined how much predictive performance  for severe COVID-19  a risk classifier built  and tested with four-fold cross validation based on these conditions could achieve and whether this could be improved using information from the rest of the hospitalisation and prescribing data.

_Findings_ -- In a logistic regression based on the age-sex distribution of the national population, the odds ratios were `r round(exp(10 * logistic.coeffs[3, 1]), 2)` for a 10-year increase in age and `r round(exp(-logistic.coeffs[2, 1]), 2)` for male sex.  In the case-control analysis, the odds ratios were 
`r table.agegr.all[1, 3]` for residence in a care home status, 
`r table.agegr.all[4, 3]` for diabetes, 
`r table.agegr.all[8, 3]` for chronic renal failure, 
`r table.agegr.all[5, 3]` for ischemic heart disease,
`r table.agegr.all[6, 3]` for other heart disease, 
`r table.agegr.all[9, 3]` for neurological disease, 
`r table.agegr.all[10, 3]` for chronic liver disease and 
`r table.agegr.all[11, 3]` for immune deficiency or suppresssion.

Overall `r freqs.all[13, 2]` of severe cases and `r freqs.all[13, 1]` of controls had at least one of these listed conditions: `r gsub("\\\\", "", table.agegr[13, 2])` versus `r gsub("\\\\", "", table.agegr[13, 1])` in those aged under 40 years.  The odds ratio for severe disease associated with encashment of at least one prescription drug in the past year was `r table.agegr.all[2, 3]` and the odds ratio associated with at least one hospital admission in the past five years was `r table.agegr.all[3, 3]`).  These associations with severe disease were present in those without any of the listed conditions, and were distributed over most BNF chapters and the associations with hospital diagnoses over most ICD-10 chapters.

With a model based on age and sex alone, the expected information for discrimination was `r round(cases.Lambda.agesex, 2)` bits.  The additional information for discrimination conditional on age and sex was `r table.densities[1, 5]` bits for a model based on demographic variables only, `r table.densities[2, 5]` bits for a model that included listed conditions, and `r table.densities[3, 5]` bits for a model that included indicator variables for hospital diagnoses and prescription encashments.

_Conclusions_ --Along with older age and male sex, severe COVID-19 disease is strongly associated with past medical history and most younger persons developing severe disease have an underlying medical condition. The comorbidities associated with increased risk are not limited to the current lists of risk conditions highlighted by public health agencies.  A risk classifier that uses all the information available in medical and prescribing records, rather than only a limited set of conditions, will more accurately discriminate between low-risk and high-risk individuals who may require shielding until the SARS-CoV-2 epidemic is over.

@@

The quantification of the risk associated with  characteristics and co-morbidities has been limited by the lack of an appropriate background population comparator  [@zhou_clinical_2020;@grasselli_baseline_2020;@guan_clinical_2020].   Two recent UK analyses have  had population comparators and these have reported  increased relative risks for a range of co-morbidities including diabetes, asthma and heart disease and for non-white ethnicity[@niedzwiedz_ethnic_2020;@the_opensafely_collaborative_opensafely_2020].   A systematic evaluation of the past medical history in those with and without severe disease has not as yet been reported.

We undertook a systematic examination of the health care record and prescribing history among all people who have developed severe COVID-19 disease in Scotland compared to the background population to understand sociodemographic and comorbidity factors associated with severe COVID-19 disease.

# Methods
## Case definition 
The Electronic Communication of Surveillance in Scotland (ECOSS) database captures all virology testing in all NHS laboratories nationally, using the Community Health Index (CHI) identifier.  The CHI number is a unique identifier for everyone in the population, and is used in all care systems in Scotland.  All individuals testing positive for nucleic acid for SARS-CoV-2 up to `r format(max(cc.all$SPECDATE, na.rm=TRUE), "%d %B %Y")` were ascertained.  Hospital admissions from the time of testing were obtained from the RAPID-24 database which is updated each day and the weekly updated Scottish Morbidity Record 01. Admissions to critical care were obtained from the Scottish Intensive Care Society and Audit Group (SICSAG) database that captures admission to all critical care units across Scotland. Death registrations up to `r format(max(cc.all$Date.Death, na.rm=TRUE), "%d %B %Y")` were obtained from linkage to the National Register of Scotland.

Severe or fatal COVID-19 was defined by a record of entering critical care in the SICSAG database, or death within 28 days of a positive nucleic acid test, regardless of the cause of death given on the death certificate.  By restricting the case definition to a group that combines those receiving critical care with those who died without receiving critical care, we ensured complete ascertainment of severe test-positive cases, whatever selection procedures may have been in place to exclude those assessed as unlikely to benefit from critical care.

@@

## Demographic data
Socioeconomic status was scored in the CHI database as the Scottish Index of Multiple Deprivation, an indicator based on postal code.  Ethnicity was assigned based on applying a name classification algorithm ONOMAP [@lakha_name_2011] to the names in the CHI database.  This classification is able to distinguish South Asian and Chinese groups with fair accuracy but does not distinguish those who identify as Black or Black-Caribbean from those who identify as White. For those cases and controls who had a past record of hospital admission or outpatient attendance in the Scottish Morbidity Record, a separate classification of ethnicity was available, assigned by self-report or by hospital staff and based on the categories used in the Census.

@@

We first extracted history of a specific list of conditions and drug classes that have been included as risk conditions for COVID-19 by public health agencies [@nhs_whos_2020].  These co-morbid conditions were assigned based on either a diagnostic code in any hospital discharge record during the last five years, or an encashed prescription of a drug for which that diagnostic code is the only indication.  The ICD-10 codes and drug codes used to define each variable are listed in the R script provided as Supplementary Material. In addition to these pre-specified groupings of codes we then undertook a systematic evaluation of the association of all ICD-10 codes and BNF codes to chapter level.  We did not specifically tabulate the conditions that have been listed as making people " clinically extremely vulnerable" such as cystic fibrosis or leukaemia  or organ transplantation as there is no clinical equipoise regarding risk in such persons and as it is expected that they will be underrepresented in the case data as they have been advised to shield themselves completely since early in the epidemic. This group also includes those with cancer in receipt of chemotherapy or radiotherapy.

## Statistical methods
To estimate the relation of cumulative incidence and mortality from COVID-19 to age and sex, logistic regression models were fitted to the proportions of cases and non-cases in the Scottish population, using the estimated population of Scotland in mid-year 2019 which are available by one-year age group up to age 90 years. For the case-control study, estimates of associations with severe COVID-19 were based on conditional logistic regression, implemented as Cox regression in the R function `survival::clogit`.  Among those cases and controls without any of the pre-defined conditions we then further examined associations of ICD-10 and BNF chapter with severe COVID-19. Where the sample of cases and controls is restricted, this will generate strata that contain no cases.  These strata will be ignored by the conditional logistic regression model as they do not contribute to the conditional likelihood.  With incidence density sampling, odds ratios in conditional logistic regression models are equivalent to rate ratios.  These odds ratios based on the conditional likelihood will differ from the unconditional odds ratios calculable from the frequencies of exposure in cases and controls [@breslow_estimation_1978]

To construct risk prediction models, we used stepwise regression alternating between forward and backward steps to maximize the AIC, implemented in the R function `stats::step`.  The performance of the risk prediction model as a classifier was examined by 4-fold cross-validation, with performance calculated over all test folds as the expected information for discrimination $\Lambda$ [@mckeigue_quantifying_2019] and (more conventionally) as the C-statistic. The rationale for using $\Lambda$ to quantify the predictive performance of a binary classifier, and especially to quantify increments in predictive performance, has been explained elsewhere [@mckeigue_quantifying_2019].  The contributions of independent predictors are additive on the scale of $\Lambda$.  In this study we have estimated the information for discrimination provided by age and sex (from a logistic regression model in the general population), and the information for discrimination provided by other risk factors conditional on age and sex (from conditional logistic regression models in this case-control study matched for age and sex).  We can add these to obtain the total information for discrimination that would be obtained with a classifier in the general population.  As described elsewhere, this can be used to estimate the proportion of those destined to develop severe or fatal disease who would be correctly classified as high risk under a shielding policy [@mckeigue_evaluation_2020].

@@

# Results
## Incidence and mortality from severe COVID-19 in the Scottish population
Figure \ref{fig:rates} shows the relationships of incidence and mortality rates to age for each sex separately.  The relationship of mortality to age is almost exactly linear on a logit scale, and the lines for males and females are almost exactly parallel.  For severe cases as defined in this study, the sex differential is narrow up to about age 50, implying that at younger ages the ratio of critical care admissions to total fatalities is higher in women than in men, but widens from about age 40 onwards.  In models including both age and sex as covariates, the odds ratio associated with a 10-year increase in age was `r round(exp(10*logistic.coeffs[3, 1]), 2)` for all severe disease and `r round(exp(10*logistic.coeffs[3, 2]), 2)` for fatal disease.  The odds ratio associated with male sex was `r round(exp(-logistic.coeffs[2, 1]), 2)` for all severe disease and `r round(exp(-logistic.coeffs[2, 2]), 2)` for fatal disease.  As shown in Table \ref{tab:agegr} almost all severe cases over aged 75 years died but most severe cases under aged 60 survived.

## Association with demographic factors
Table \ref{tab:demog} shows univariate associations of demographic factors with severe disease.  Higher risk of severe disease was associated with socioeconomic deprivation and with residence in a care home.  In comparison with those classified as White, South Asians and other non-White ethnic minorities were at lower risk of severe disease.  Table \ref{tab:ethnicsmr} shows a separate analysis restricted to those with ethnicity recorded in hospital records, which allows those who identify as Black to be distinguished from those who identify as White.  Although the numbers are small, there is no evidence that the Black minority is at higher risk of severe disease.

@@

As shown  in the age band specific univariate logistic regression models in tables \ref{tab:listed.conditions.lt60} to \ref{tab:listed.conditions.ge75}.  Here we collapsed the two younger age bands to <60 years.  All of the listed conditions were significantly associated with severe disease  in each age band. Diabetes was associated with a 6.6 fold rate of severe COVID-19 under age 60 years.  Of note is that "other heart disease" was more  strongly associated with severe COVID-19 than ischaemic heart disease.

Tables \ref{tab:listed.conditions.lt60} to \ref{tab:listed.conditions.ge75}  also show  multivariate rate ratios by age group. The key point of this  analysis is to demonstrate that at all ages there was a very strong  and highly significant  association of having had any prescription in the last year  and of having had any admission in the past five years even adjusted for all the listed conditions and care home residence.

## Systematic analysis of diagnoses associated with severe disease
The approximately two-fold risk of with severe COVID-19 associated with prior hospital admission independently of other conditions and factors was further explored by systematically testing for association of hospitalisations at each ICD-10 chapter level with severe COVID-19, restricted to those without any of the listed conditions.  These results are shown in Table \ref{tab:conditions}.  In univariate analyses, almost all ICD chapters, with the exception of Chapters 8 (ear) and Chapter 15 (pregnancy) were associated with increased risk of severe disease.  The strongest associations were seen with neoplasms and with eye disease. Table \ref{tab:icdsubchapters} extracts univariate associations with ICD subchapters where the p-value is <0.001 and where there are at least 50 cases and controls with a diagnosis in this subchapter.  The key point of this table shows that there are many subchapter diagnoses that are significantly more frequent in severe COVID-19 cases than controls.  Of particular note is that prior hospitalisations with infections, pneumonia and acute respiratory diseases, circulatory disorders other than heart disease, including cerebrovascular accidents.  Without filtering on the aforementioned criteria there are of course additional differences in counts of rarer diagnoses in cases and controls are apparent.

## Associations of prescribed drugs with severe disease 
Tables \ref{tab:listed.conditions.lt60} to \ref{tab:listed.conditions.ge75} show that the strongest risk factor for severe disease, apart from residence in a care home, is the encashment of any prescription in the last year.  The univariate rate ratio associated with this variable varies from `r tables.agegr[[1]][2, 3]` in those aged under 60 years to `r tables.agegr[[3]][2, 3]` in those aged 75 years and over.  In a multivariate analysis adjusting for care home residence, any hospital admission and listed conditions, these rate ratios were reduced to `r tables.agegr[[1]][2, 5]` and  `r tables.agegr[[3]][2, 5]`   About one third of those over 75 had not encashed a prescription in the previous year.

@@

## Construction of a multivariate risk prediction model 
To obtain a summary metric of the additional information there is in the hospitalisation and prescribing data predictive models were constructed from two sets of variables: a baseline set consisting only of care home residence plus the listed conditions, and an extended set that augmented the baseline set with an indicator variable for ICD-10 chapter level admissions, and indicator variables for prescriptions in each of the BNF chapters.

For each of variable set, a stepwise regression procedure was carried out using alternating forward-backward selection. The variables retained with each variable set are shown in Table \ref{tab:stepwise} . The predictive performance of the model chosen by stepwise regression was estimated by `r nfold`-fold cross-validation.  Observed and predicted case status were compared over all test folds.  Table \ref{tab:xvalid} shows that using the extended set increased the C-statistic from `r table.densities[2, 3]` to `r table.densities[3, 3]` and the expected information for discrimination from `r table.densities[2, 5]` bits to `r table.densities[3, 5] ` bits.

@@

Figure \ref{fig:wevid} shows the class-conditional distributions of the weight of evidence $W$ favouring case over control status from the model based on the extended variable set.  The model-based distributions have been adjusted slightly to make them mathematically consistent as described elsewhere [@mckeigue_quantifying_2019].  This shows, as expected for a multifactorial classifier, that the distributions are approximately Gaussian: there is no clear divide between high-risk and low-risk individuals of the same age and sex. Figure \ref{fig:roc}, which shows the corresponding receiver operating characteristic (ROC) curve, encodes exactly the same information as Figure \ref{fig:wevid} in a different form.  The likelihood ratio $2^W$ at any given sensitivity and specificity is the gradient of the model-based ROC curve.

# Discussion  
### Sociodemographic features
In this analysis of sociodemographic factors and past medical history we found a large number of factors to contribute to risk of severe COVID-19 disease.  Apart from the very steep rise in risk with age, severe disease is more likely in men, and those who are more socioeconomically disadvantaged.  Regarding sex differences we noted that the ratio of non-fatal cases who receive critical care to fatal cases is lower in females than in males in the 50-70 year age range.  This may reflect either an underlying divergence in case fatality, or that women are less likely than men to be transferred to critical care if they have a good chance of surviving without critical care.  The effect of matching on general practice which will partly match for socioeconomic disadvantage will reduce the precision of the effect of the socioeconomic indicator (SIMD) in our analysis but will not bias it in either direction.  Residence in a care home was associated with a 15-fold increased risk of severe COVID-19 in this age matched analysis and this was reduced to eight-fold adjusted for the listed conditions. This eight-fold risk likely reflects both the spread of the epidemic in care homes and some residual confounding by frailty.

In contrast to reports from the US and England [@the_opensafely_collaborative_opensafely_2020] we found no evidence of increased risk in non-White minorities. This suggests that ethnic differences reported in other countries are not genetically driven but reflect other social factors that differ between regions in their association with ethnicity.  We note that the proportion of health care workers that are of non-British nationality varies across the UK from [23% in London to 7% in Scotland](https://www.ons.gov.uk/peoplepopulationandcommunity/populationandmigration/internationalmigration/articles/internationalmigrationandthehealthcareworkforce/2019-08-15).  In a recent report based on the UK Biobank cohort, almost half of the fourfold risk ratio for hospitalised COVID associated with Black ethnicity was accounted for by health care worker status, with other social variables attenuating the risk further[@niedzwiedz_ethnic_2020; @ho_modifiable_2020].  In the OpenSAFELY prospective cohort study using medical records in England compared to those recorded as White, Black and Asian people were at 1.7 fold and 1.6 fold risk of death respectively adjusted for comorbidities and socioeconomic status, However as acknowledged by the authors, data on health care worker status and other social factors would be needed to explore this further. We are in the process of obtaining more detailed data on health care worker status in Scotland.

### Co-morbidities
We have confirmed the increased risks associated with the listed conditions. These were based on the list of " moderate risk " conditions highlighted by the NHS and other agencies [@nhs_whos_2020].

The risk associated with these conditions various by age - for example the odds ratio associated with diabetes is much higher at younger ages. The 2.7 fold risk in all ages combined is commensurate with that reported in UK Biobank and in the OpenSAFELY studies.  We confirm the higher risk with asthma and chronic lung disease and liver disease reported in these and earlier studies.  Of note other heart disease is more strongly associated than ischaemic heart disease.  This category includes many conditions such as atrial fibrillation, cardiomyopathies and heart failure.  However overall 26% of cases did not have any of these conditions.  Among cases and controls without these conditions not surprisingly neoplasms were associated with severe COVID-19; we had omitted it from the pre-specified list as many people under current chemotherapy and radiotherapy are included in the shielding at present.  We have not focussed on evaluating the risk associated with conditions for which shielding is recommended, as the observed risk will depend on the adequacy of shielding.  A further systematic evaluation of past hospitalisation history did not reveal a sparse set of conditions underlying severe disease but rather very many associated diagnoses. We note however that past hospitalisation for influenza and respiratory tract infections is an important risk factor.

>>>>

_Background_-- Why some people infected with SARS-CoV-2 develop severe COVID-19 disease and others do not is poorly understood.  The objectives of this study were to identify risk factors for severe disease and to lay the basis for risk stratification based on demographic data and health records.

_Methods_ -- The design was a matched case-control study. Test-positive cases were ascertained as all those with a positive nucleic acid test for SARS-CoV-2 in the national database and from national mortality data.  Those who had been admitted to intensive care or had died within 28 days of a positive test were classified as severe cases.  For each case, ten controls matched for sex, age and primary care practice were selected from the population register.  For cases and controls we extracted all diagnostic codes from the past five years of hospital discharge records, all drug codes from encashed prescriptions issued in primary care care during the past year and linked to the national diabetes register.  Rate ratios for associations with severe COVID-19 were estimated by conditional logistic regression.  The performance of a predictive model for severe COVID-19 was estimated by cross-validation.

_Findings_ -- In a logistic regression based on the age-sex distribution of the national population, the odds ratios were `r round(exp(10 * logistic.coeffs[3, 1]), 2)` for a 10-year increase in age and `r round(exp(-logistic.coeffs[2, 1]), 2)` for male sex.  In the case-control analysis, the strongest risk factor was residence in a care home, with rate ratio (95% CI) `r table.agegr.all[1, 3]` .  Rate ratios (95% CIs) for conditions previously designated as conferring high risk were  
`r table.agegr.all[4, 3]` for diabetes, 
`r table.agegr.all[8, 3]` for chronic renal failure, 
`r table.agegr.all[5, 3]` for ischemic heart disease,
`r table.agegr.all[6, 3]` for other heart disease, 
`r table.agegr.all[9, 3]` for neurological disease, 
`r table.agegr.all[10, 3]` for chronic liver disease and 
`r table.agegr.all[11, 3]` for immune deficiency or suppression.

Overall `r freqs.all[13, 2]` of severe cases and `r freqs.all[13, 1]` of controls had at least one of these listed conditions.  In those aged under 40 years the frequencies of any listed condition were `r gsub("\\\\", "", table.agegr[13, 2])` in cases and `r gsub("\\\\", "", table.agegr[13, 1])` in controls.  The odds ratio for severe disease associated with encashment of at least one prescription in the past year was `r table.agegr.all[2, 3]` and the odds ratio associated with at least one hospital admission in the past five years was `r table.agegr.all[3, 3]`).  These associations remained after excluding those with any of the listed conditions.  The associations with hospital diagnoses were distributed over most chapters of the ICD-10 and the associations with prescribed drugs were distributed over most chapters of the British National Formulary.

With a model based on age and sex alone, the expected information for discrimination was `r round(cases.Lambda.agesex, 2)` bits.  The additional information for discrimination conditional on age and sex was `r table.densities[1, 5]` bits for a model based on demographic variables only, `r table.densities[2, 5]` bits for a model that included the listed conditions, and `r table.densities[3, 5]` bits for a model that included indicator variables for hospital diagnoses and prescription encashments.

_Conclusions_ --Along with older age and male sex, severe COVID-19 disease is strongly associated with past medical history; most younger persons developing severe disease have an underlying medical condition. The comorbidities associated with increased risk are not limited to the lists of risk conditions designated by public health agencies.  A risk classifier that uses all the information available in medical and prescribing records, rather than only a limited set of conditions, will more accurately discriminate between low-risk and high-risk individuals who may require shielding until the SARS-CoV-2 epidemic is over.

@@

Quantification of the risk associated with characteristics and co-morbidities has been limited by the lack of comparisons with the background population [@zhou_clinical_2020;@grasselli_baseline_2020;@guan_clinical_2020].  Two recent studies in the UK have included population comparators and have reported associations with co-morbidities including diabetes, asthma and heart disease. [@niedzwiedz_ethnic_2020; @the_opensafely_collaborative_opensafely_2020].  A systematic comparison of medical records in those with and without severe disease has not yet been reported.

We report a systematic examination of the health care record and prescribing history among all people who have developed severe COVID-19 disease in Scotland compared to the general population. The objectives were to identify risk factors and to lay the basis for risk stratification based on a predictive model.

# Methods
## Case definition 
The Electronic Communication of Surveillance in Scotland (ECOSS) database captures all virology testing in all NHS laboratories nationally, using the Community Health Index (CHI) identifier.  The CHI number is a unique identifier used in all care systems in Scotland.  All individuals testing positive for nucleic acid for SARS-CoV-2 up to `r format(max(cc.all$SPECDATE, na.rm=TRUE), "%d %B %Y")` were ascertained.  Hospital admissions from the time of testing were obtained from the RAPID-24 database which is updated each day and from the Scottish Morbidity Record 01 which is updated weekly.  Admissions to critical care were obtained from the Scottish Intensive Care Society and Audit Group (SICSAG) database that covers all critical care units in Scotland. Death registrations up to `r format(max(cc.all$Date.Death, na.rm=TRUE), "%d %B %Y")` were obtained from linkage to the National Register of Scotland.

Severe or fatal COVID-19 was defined by a record of entering critical care in the SICSAG database, or death within 28 days of a positive nucleic acid test, regardless of the cause of death given on the death certificate.  By restricting the case definition to cases that were fatal or received critical care, we ensured complete ascertainment of all test-positive cases that were severe enough to have been fatal without critical care, irrespective of any selection that may have excluded those assessed as unlikely to benefit from critical care.

@@

## Demographic data
Socioeconomic status was scored in the CHI database as the Scottish Index of Multiple Deprivation, an indicator based on postal code.  Ethnicity was assigned based on applying a name classification algorithm (ONOMAP) [@lakha_name_2011] to the names in the CHI database.  This classification is able to distinguish South Asians, the largest non-White minority in Scotland, from those who identify as White with fair accuracy but misclassifies most of those who identify as Black or Black-Caribbean (a much smaller minority in Scotland) as White. <!--- For cases and controls who had a past record of hospital admission or outpatient attendance in the Scottish Morbidity Record, a separate classification of ethnicity was available, assigned by self-report or by hospital staff and based on the categories used in the Census. --->

@@

We began by scoring a specific list of conditions that have been designated as risk conditions for COVID-19 by public health agencies [@nhs_whos_2020].  These listed conditions were assigned based on diagnostic codes in any hospital discharge record during the last five years, or encashed prescription of a drug for which the only indications are in that group of diagnostic codes. The ICD-10 codes and drug codes used to define each variable are listed in the R script provided as Supplementary Material. In addition to these pre-specified groupings of codes we tabulated associations with all ICD-10 codes and BNF codes to chapter level.  A
separate list of conditions has been used to designate "clinically extremely vulnerable" who have been advised to shield themselves completely since early in the epidemic: this list includes solid organ transplant recipients, people receiving chemotherapy for cancer, and people with cystic fibrosis or leukaemia.  We did not separately tabulate these conditions as we expected these individuals to be underrepresented among cases if shielding was adequate.

## Statistical methods
To estimate the relation of cumulative incidence and mortality from COVID-19 to age and sex, logistic regression models were fitted to the proportions of cases and non-cases in the Scottish population, using the estimated population of Scotland in mid-year 2019 which are available by one-year age group up to age 90 years. For the case-control study, estimates of associations with severe COVID-19 were based on conditional logistic regression, implemented as Cox regression in the R function `survival::clogit`.  Among those cases and controls without any of the pre-defined conditions we then further examined associations of ICD-10 and BNF chapter with severe COVID-19. Where the sample of cases and controls is restricted, this will generate strata that contain no cases.  These strata will be ignored by the conditional logistic regression model as they do not contribute to the conditional likelihood.  With incidence density sampling, the odds ratios in conditional logistic regression models are equivalent to rate ratios.  These odds ratios based on the conditional likelihood will differ from the unconditional odds ratios calculable from the frequencies of exposure in cases and controls [@breslow_estimation_1978]

To construct risk prediction models, we used stepwise regression alternating between forward and backward steps to maximize the AIC, implemented in the R function `stats::step`.  The performance of the risk prediction model as a classifier was examined by 4-fold cross-validation, with performance calculated over all test folds as the expected information for discrimination $\Lambda$ [@mckeigue_quantifying_2019] and (more conventionally) as the C-statistic. The rationale for using $\Lambda$ to quantify the predictive performance of a binary classifier, and especially to quantify increments in predictive performance, has been explained elsewhere [@mckeigue_quantifying_2019].  A key advantage of using $\Lambda$ is that the contributions of independent predictors are additive on the scale of $\Lambda$.  The class-conditional distributions of weights of evidence in cases and controls encode exactly the same information as the receiver operating characteristic curve, when these distributions are adjusted for mathematical consistency as described elsewhere [@mckeigue_quantifying_2019] to obtain the model-based distributions.  The likelihood ratio at any given sensitivity and specificity is the gradient of the model-based ROC curve.

In this study we have estimated the information for discrimination provided by age and sex (from a logistic regression model in the general population), and the information for discrimination provided by other risk factors conditional on age and sex (from conditional logistic regression models in this case-control study matched for age and sex).  We can add these to obtain the total information for discrimination that would be obtained with a classifier in the general population.  As described elsewhere, this can be used to estimate the proportion of those destined to develop severe or fatal disease who would be correctly classified as high risk under a shielding policy [@mckeigue_evaluation_2020].

@@

# Results
## Incidence and mortality from severe COVID-19 in the Scottish population
Figure \ref{fig:rates} shows the relationships of incidence and mortality rates to age for each sex separately.  The relationship of mortality to age is almost exactly linear on a logit scale, and the lines for males and females are almost exactly parallel.  For severe cases as defined in this study, the sex differential is narrow up to about age 50 but widens between ages 50 and 70 years.  This implies that at younger ages the ratio of critical care admissions to total fatalities is higher in women than in men, but that at later ages the ratio of critical care admissions to total fatalities is higher in men.  In models that included  age and sex as covariates, the odds ratio associated with a 10-year increase in age was `r round(exp(10*logistic.coeffs[3, 1]), 2)` for all severe disease and `r round(exp(10*logistic.coeffs[3, 2]), 2)` for fatal disease.  The odds ratio associated with male sex was `r round(exp(-logistic.coeffs[2, 1]), 2)` for all severe disease and `r round(exp(-logistic.coeffs[2, 2]), 2)` for fatal disease.  As shown in Table \ref{tab:agegr} almost all severe cases over aged 75 years died but most severe cases under aged 60 survived.

## Association with demographic factors
Table \ref{tab:demog} shows univariate associations of demographic factors with severe disease.  Residence in a care home was by far the strongest risk factor for severe disease.  Higher risk of severe disease was also associated with socioeconomic deprivation.  In comparison with those classified as White, South Asians and other non-White ethnic minorities were at lower risk of severe disease.  Table \ref{tab:ethnicsmr} shows a separate analysis restricted to those with ethnicity recorded in hospital records, which allows those who identify as Black to be distinguished from those who identify as White.  The numbers of cases and controls classified as Black are too small for the risk in this group to be estimated reliably.

@@

Tables \ref{tab:listed.conditions.lt60} to \ref{tab:listed.conditions.ge75} examine these associations by age group, with the 0-39 and 40-59 age bands combined.  All listed conditions were associated with severe disease in each age band.  Diabetes was associated with a 6.6 fold rate of severe COVID-19 under age 60 years.  The associations with severe COVID-19 were stronger for the broad category "other heart disease" than for ischaemic heart disease.

Tables \ref{tab:listed.conditions.lt60} to \ref{tab:listed.conditions.ge75} also show multivariate rate ratios by age group. These table show that even after adjusting for care home residence and listed conditions, encashment of any prescription in the past year and any admission to hospital in the past five years were strongly and independently associated with severe disease.

## Systematic analysis of diagnoses associated with severe disease
The approximately two-fold risk of with severe COVID-19 associated with prior hospital admission independently of other conditions and factors was examined further by testing for association of hospitalisations at each ICD-10 chapter level with severe COVID-19, restricted to those without any of the listed conditions.  These results are shown in Table \ref{tab:conditions}.  In univariate analyses, almost all ICD chapters, with the exception of Chapters 8 (ear) and Chapter 15 (pregnancy) were associated with increased risk of severe disease.  The strongest residual associations were seen with neoplasms. Table \ref{tab:icdsubchapters} extracts univariate associations with ICD subchapters where the p-value is <0.001 and where there are at least 50 cases and controls with a diagnosis in this subchapter.  This table shows that many subchapter diagnoses are associated with markedly higher risk of severe COVID-19.  We note that past hospital diagnoses of infections, pneumonia and acute respiratory diseases are strongly associated with severe COVID-19.  Associations with cardiovascular disease are not limited to heart disease but are also present for stroke and other circulatory disorders that are not designated as risk conditions.

## Associations of prescribed drugs with severe disease 
Tables \ref{tab:listed.conditions.lt60} to \ref{tab:listed.conditions.ge75} show that the strongest risk factor for severe disease, apart from residence in a care home, is the encashment of at least one prescription in the last year.  The univariate rate ratio associated with this variable varies from `r tables.agegr[[1]][2, 3]` in those aged under 60 years to `r tables.agegr[[3]][2, 3]` in those aged 75 years and over.  In a multivariate analysis adjusting for care home residence, any hospital admission and listed conditions, these rate ratios were reduced to `r tables.agegr[[1]][2, 5]` and `r tables.agegr[[3]][2, 5]` About one third of those over 75 had not encashed a prescription in the previous year.

@@

## Construction of a multivariate risk prediction model 
To evaluate the incremental contributions of the listed conditions, and residual information in hospitalisation and prescription records data after assigning these conditions, predictive models were constructed from three sets of variables: a baseline set consisting only of demographic variables, a set that included indicator variables for each listed condition, and an extended set that included demographic + listed conditions + indicator variables for hospital diagnoses in each ICD-10 chapter and prescriptions in each BNF chapter.

For each variable set, a stepwise regression procedure was carried out using alternating forward-backward selection. The variables retained with each variable set are shown in Table \ref{tab:stepwise} . The predictive performance of the model chosen by stepwise regression was estimated by `r nfold`-fold cross-validation.  Observed and predicted case status were compared over all test folds.  Table \ref{tab:xvalid} shows that using the extended set increased the C-statistic from `r table.densities[2, 3]` to `r table.densities[3, 3]` and the expected information for discrimination from `r table.densities[2, 5]` bits to `r table.densities[3, 5] ` bits.

@@

Figure \ref{fig:wevid} shows the class-conditional distributions of the weight of evidence $W$ favouring case over control status from the model based on the extended variable set.  The model-based distributions have been adjusted slightly to make them mathematically consistent as described elsewhere [@mckeigue_quantifying_2019].  This shows, as expected for a multifactorial classifier, that the distributions are approximately Gaussian: there is no clear divide between high-risk and low-risk individuals of the same age and sex. Figure \ref{fig:roc} shows the corresponding receiver operating characteristic (ROC) curve, which encodes the same information as Figure \ref{fig:wevid} but does not allow the separation of cases and controls by the risk classifier to be easily visualized.

# Discussion  
### Sociodemographic features
In this analysis of sociodemographic factors and past medical history we found a large number of factors to contribute to risk of severe COVID-19 disease.  Apart from the very steep rise in risk with age, severe disease is more likely in men, and those who are more socioeconomically disadvantaged.  Regarding sex differences we noted that the ratio of non-fatal cases who receive critical care to fatal cases is lower in females than in males in the 50-70 year age range.  This may reflect either an underlying divergence in case fatality, or that women are less likely than men to be transferred to critical care if they have a good chance of surviving without critical care.  The effect of matching on general practice which will partly match for socioeconomic disadvantage will reduce the precision of the effect of the socioeconomic indicator (SIMD) in our analysis but will not bias it in either direction.  Residence in a care home was associated with a 15-fold increased risk of severe COVID-19 in this age matched analysis and this was reduced to eight-fold by adjustment for listed conditions. This eight-fold risk is likely to reflect both the spread of the epidemic in care homes and residual confounding by frailty.

In contrast to reports from England [@the_opensafely_collaborative_opensafely_2020] we found that people of South Asian ethnicity, and other non-White minorities, are at lower risk than Whites. This suggests that socioeconomic rather than genetic factors underlie at least some of the associations of risk with ethnicity reported in England.  In the OpenSAFELY cohort, those recorded as Black and Asian people were at 1.7 fold and 1.6 fold risk of death respectively compared with those recorded as White, after adjustment for comorbidities and socioeconomic status, However in a recent report based on the UK Biobank cohort, almost half of the fourfold risk ratio for hospitalised COVID associated with Black ethnicity was accounted for by health care worker status, with other social variables attenuating the risk ratio further [@niedzwiedz_ethnic_2020; @ho_modifiable_2020].  We note that the proportion of health care workers that are of non-British nationality varies across the UK from [23% in London to 7% in Scotland](https://www.ons.gov.uk/peoplepopulationandcommunity/populationandmigration/internationalmigration/articles/internationalmigrationandthehealthcareworkforce/2019-08-15).

### Co-morbidities
We have confirmed that the " moderate risk " conditions designated by the NHS and other agencies [@nhs_whos_2020] are associated with increased risk of severe COVID-19.  However the rate ratios associated with these conditions vary with age - for example the rate ratio associated with diabetes is higher at younger ages. The 2.7 fold rate ratio in all ages combined is similar to that reported in UK Biobank and in the OpenSAFELY studies.  We confirm the higher risk with asthma and chronic lung disease and liver disease reported in these and earlier studies.  Of note other heart disease is more strongly associated than ischaemic heart disease.  This category includes conditions such as atrial fibrillation, cardiomyopathies and heart failure.  However overall 26% of cases did not have any of these listed conditions.  Among cases and controls without these conditions not surprisingly neoplasms were associated with severe COVID-19; we had omitted it from the pre-specified list as in the current dataset we cannot separately identify those who are currently receiving chemotherapy or radiotherapy for whom shielding is advised.  We have not attempted to estimate the risk associated with these conditions for which shielding is recommended, as the observed risk will depend on the adequacy of shielding rather than on the risk to those exposed to the epidemic.  A further systematic evaluation of past hospitalisation history did not reveal a sparse set of conditions underlying severe disease but rather many associated diagnoses.  Past hospital admission for influenza or respiratory tract infection, though not usually designated as a high-risk condition, is an important and easily-scored risk factor for severe COVID-19.
